---
title: "Integrating RNA, phospho-proteomic, and proteomic data with PLS in Mixomics"
output: html_document
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=TRUE, message=FALSE, warning=FALSE, result='hide'}
pkg=c("mixOmics", "dplyr")

lapply(pkg, require, character.only = TRUE)


setwd("/Users/jason/Documents/git_repos/cancer_omics")
```


Given proteomics, rna, and phospho-proteomics datda, can you predict phosphorylation response to EGF treatment? 

We have initial state for each of these 3 data typles. The goal is to predict the phosphorylation of MEKK4 and IdU after 1 hour of EGF treatment


Load proteomics and RNA data:
```{r load proteomics/rna data}
rna=read.csv("/Users/jason/Documents/git_repos/cancer_omics/rnaseqexononly_marcotte.csv")
prot= read.csv("/Users/jason/Documents/git_repos/cancer_omics/proteomics_consolidated.csv")

#set cell_line to row_names
rownames(rna)= rna$X
rownames(prot)= prot$X

rna=select(rna, -"X")
prot=select(prot, -"X")


#order by rownames
rna=rna[order(row.names(rna)),]
prot=prot[order(row.names(prot)),]

```


Load phospho-proteomics data:
```{r load phosphorylation data}
phosphoprot=read.csv("/Users/jason/Documents/git_repos/cancer_omics/median_phospho_data.csv")
phosphoprot =phosphoprot[ order(phosphoprot$cell_line), ] #sort by cell_line name
#basal=phospho[phospho$treatment=="full",]

#control/baseline phospho data used for prediction
phospho=filter(phosphoprot, treatment=="EGF", time==0) #baseline phospho data
rownames(phospho)= phospho$cell_line #set cell_line to rowname
phospho=select(phospho, -c("treatment","time","cell_line","p.MKK4","p.GSK3b")) #drop unneeded & predictor columns


##predictor column
Y=filter(phosphoprot, treatment=="EGF", time==60)  #predicted column, EGF phoshorylation result 1hr after treatment
rownames(Y)= Y$cell_line

Y= select(Y,c('p.MKK4','p.GSK3b')) #drop other columns


#comparing repsonses for each phosphoprotein
boxplot(phospho, use.columns=TRUE, xlab="phospho-protein", ylab="phosphorylation level", main="Phoshporylation distributions") #skip non-numeric columns



```


Train-test split on data. Creating input X matrix and dependent matrix Y

```{r train-test split }
#final list has 59 cell lines, split off 10% (6 lines from test set validation)
cell_inter=intersect(rownames(rna), rownames(prot))
cell_inter= intersect(cell_inter, rownames(phospho))

incommon=length(cell_inter)
#train/test: 90/10 split based on cell lines
train= head(cell_inter, incommon-6)
test= tail(cell_inter, 6)


rna_train=rna[rownames(rna) %in% train,]  %>% as.matrix()
prot_train=prot[rownames(prot) %in% train,]  %>% as.matrix()
phos_train=phospho[rownames(phospho) %in% train,] %>% as.matrix()

Ytrain=Y[rownames(Y) %in% train,] %>% as.matrix() #2 columns


rna_test=rna[rownames(rna) %in% test,] #%>% as.matrix()
prot_test=prot[rownames(prot) %in% test,] # %>% as.matrix()
phos_test=phospho[rownames(phospho) %in% test,] # %>% as.matrix()
Ytest=Y[rownames(Y) %in% test,] %>% as.matrix()


#mixomics expects list of matrices (with each matrix containing a datatype)
X <- list(RNA = rna_train ,
          phospho = phos_train,
          protein = prot_train)


```



Fit regression model. 
```{r fitting}
#reduce complexity, remove if you have low sample number
#keepx: number of (top) variables to select from the first two components of PCA
#list.keepX <- list(RNA = c(20, 20), phospho = c(8,8), protein = c(5, 5))

#features are centered and scaled by default
MyResult.diablo <- block.pls(X, Ytrain)
```



```{r extracting additional info from pls object}
loadings1=selectVar(MyResult.diablo, comp = 1) #loadings for component 1
loadings2=selectVar(MyResult.diablo, comp = 2) #loadings for component 2

#loadings$RNA$value # will have loadinpgs for RNA block

#raw data can be pulled from :
#MyResult.diablo$X


#number of components used for each block:
#MyResult.diablo$ncomp

#variates
#MyResult.diablo$variates

#weighting on each components
#MyResult.diablo$weights

```




Plot where individual cell lines fit on new components in each data type.
```{r sample plot}
#reduce complexity, remove if you have low sample number
#keepx: number of (top) variables to select from the first two components of PCA
#list.keepX <- list(RNA = c(20, 20), phospho = c(8,8), protein = c(5, 5))

#features are centered and scaled by default
MyResult.diablo <- block.pls(X, Ytrain)

plotIndiv(MyResult.diablo, ind.names = FALSE)


```


Documentation on how the "predict" function works
https://rdrr.io/cran/mixOmics/man/predict.html

Generating predictions based on the pls model and evaluating the predictions.

```{r generate predictions and evaluate}
#assembling test data
Xtest <- list(RNA = rna_test ,
          phospho = phos_test,
          protein = prot_test)

#generating predictions and extracting data
ypredict=predict(MyResult.diablo, newdata = Xtest)


#missing values in proteomic test data lead to NaN prediction from that block
#instead of averaging from all 3 datatypes, average from the RNA and phospho block

#grab dim2 columns and average
avg_pred=(ypredict$predict$RNA[,,2] +ypredict$predict$phospho[,,2])/2

#phosphorylation predictions aftr 1hr EGF treatment for 6 cell lines"
print(avg_pred)



#calculate RMSE
rmse=colMeans((avg_pred- Ytest)^2)
print(rmse)

```



Weighting of each feature in each loading.
```{r variable plot, warning=FALSE}
plotVar(MyResult.diablo, overlap= F) # too many features to visualize

```


Top features for each loading in component 1 and 2
```{r visualize component 1 loadings}
#component 1 loading
plotLoadings(MyResult.diablo,comp=1, contrib = "max")
```


```{r visualize component 2 loadings}
#component 2 loading
plotLoadings(MyResult.diablo,comp=2, contrib = "max")
```


How do the top loadings match expectations? The phospho-proteomic data has the largest loadings by magnitude and largest proportion of explained variance. The top loadings for the phospho-proteome align with expectations for both MEKK4 and p.GSK3b. 

- MEKK4 is mitogen activated protein kinase, and is expected to be phosphoylated with mitogen treatment (such as EGF). Within the network of MEKK4 signaling are: other MAPK/MEKK/ERK signaling and TGF-beta/SMAD signaling.
We find the components of the loadings to be significant and in the same direction as p.MKK4. 

```{r p.Mekk4 in component1 }
#correlation with downregulation of p.mkk4
head(loadings1$phospho$value,15)
```



With component 2, we can better see correlations with p-GSK3b. We see a similar trend with known targets beta catenin and 4EBP1.
```{r p.gsk3b in component2 }
#correlation with downregulation of p.mkk4
head(loadings2$phospho$value,15)
```





Resources:

https://mixomicsteam.github.io/Bookdown/
https://www.bioconductor.org/packages/release/bioc/vignettes/mixOmics/inst/doc/vignette.html
man page: https://www.bioconductor.org/packages/devel/bioc/manuals/mixOmics/man/mixOmics.pdf
https://rstudio-pubs-static.s3.amazonaws.com/376158_5a59d491abfd4266acaa05773bc1751a.html